\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-07-18 paper/ms.tex
more typos and refs

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.95\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
    $S(d_j)/S(0) = 100$.
    (middle) The corresponding autocorrelation functions with the same colors.
    \response{(right) Three realizations in arbitrary units from each model in
    the same colors.}
    \figurelabel{sho}}
\end{center}
\end{figure}

@@

In any application of \celerite~--~or any other GP model, for that
matter~--~we must choose a kernel.
A detailed discussion of model selection is beyond the scope of this paper and
the interested reader is directed to Chapter~5 of \citet{Rasmussen:2006} an
introduction in the context of GPs, but we include a summary and some
suggestions specific to \celerite\ here.

@@

Sometimes, it is not possible to choose a physically motivated kernel when we
use a GP as an effective model.
In these cases, we recommend starting with a mixture of stochastically-driven
damped SHOs (as discussed in \sect{sho}) and select the number of oscillators
using Bayesian model comparison, cross-validation, or another model selection
technique.
Despite the fact that it is less flexible than a general \celeriteterm\
kernel, we recommend the SHO model in most cases because it is 1-time mean
square differentiable
\begin{eqnarray}
\left.\frac{\dd k_\mathrm{SHO}(\tau)}{\dd \tau}\right|_{\tau=0} = 0 \quad,
\end{eqnarray}
giving smoother functions than the general \celeriteterm\ model that is only
mean square continuous.

@@

In this first example, we simulate a dataset using a known \celeriteterm\
process and fit it with \celerite\ to show that valid inferences can be made
in this idealized case.
\response{We simulated a small ($N = \exampleindata$) dataset using a GP model
with a SHO kernel (\eqalt{sho-kernel}) with parameters $S_0 =
1\,\mathrm{ppm}^2$, $\omega_0 = e^2\,\mathrm{day}^{-1}$, and $Q = e^2$, where
the units are arbitrarily chosen to simplify the discussion.
We add further white noise with the amplitude $\sigma_n = 2.5\, \mathrm{ppm}$
to each data point.
The simulated data are plotted in the top left panel of
\Figure{simulated-correct}.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/simulated/correct.pdf}
\caption{(top left) A simulated dataset (black error bars), and MAP
    model (blue contours).
    (bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
    (right) The inferred PSD~--~the blue contours encompass 68\% of the
    posterior mass~--~compared to the true PSD (dashed black line).
    \figurelabel{simulated-correct}}
\end{center}
\end{figure}

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/simulated/wrong-qpo.pdf}
\caption{(top left) A simulated dataset (black error bars), and MAP
    model (blue contours).
    (bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
    (right) The inferred period of the process. The true period is indicated
    by the vertical orange line, the posterior inference using the correct
    model is shown as the blue dashed histogram, and the inference made using
    the ``wrong'' effective model is shown as the black histogram.
    \figurelabel{simulated-wrong}}
\end{center}
\end{figure}

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/simulated/bic.pdf}
\caption{\response{
    The Bayesian information criterion (BIC) evaluated for the data from
    \Figure{simulated-wrong} and different kernel choices.
    The $x$-axis shows the number of terms included in each model.
    The blue circles show models where the terms have been summed and the
    orange squares indicate that the model is a product of terms.
    In the product models, the parameters $S_0$ and $Q$ are held fixed at $1$
    and $1/\sqrt{2}$ respectively for all but the first term.
    The dashed black line shows the minium value of the BIC and this
    corresponds to the model chosen in the text.}
    \figurelabel{bic}}
\end{center}
\end{figure}

@@

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P_\mathrm{rot}} \right) \right)
\end{eqnarray}
where $P_\mathrm{rot}$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We construct a simple \celeriteterm\ covariance function with similar
properties as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{B}{2+C}\,e^{-\tau/L}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{rot}}\right) + (1 + C)
\right]
\end{eqnarray}
for $B>0$, $C>0$, and $L>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a significant gain in computational efficiency.

@@

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln
P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
samples.
These samples give a posterior constraint on the period of $P_\mathrm{rot} =
\rotationperiod$ days and the marginalized posterior distribution for $P$ is
shown in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/rotation/rotation.pdf}
\caption{Inferred constraints on a quasiperiodic GP model using the covariance
    function in \eq{rot-kernel} and two quarters of \kepler\ data.
(top left) The \kepler\ data (black points) and the MAP model
    prediction (blue curve) for a 60~day subset of the data used.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
(right) The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
    figure shows the posterior distribution marginalized over all other
    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
    published rotation period made using the full
    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
    \figurelabel{rotation}}
\end{center}
\end{figure}

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/astero/astero.pdf}
\caption{
\response{
(top) The \kepler\ data (black points) and the MAP model
    prediction (blue curve) for a 10~day subset of the month-long dataset that
    was used for the fit.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom) The residuals between the mean predictive model and the data
    shown in the top figure.
}
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{figures/transit/transit-ml.pdf}
    \caption{\emph{(top)} A month-long segment of \kepler\ light curve for
    KIC~1430163 with a synthetic transit model injected (black points) and the
    MAP model for the stellar variability
    (blue line).
    \emph{(bottom)} The MAP ``de-trending'' of the data in
    the top panel.
    In this panel, the MAP model for the stellar variability
    has been subtracted to leave only the transits.
    The de-trended light curve is shown by black error bars and the MAP
    transit model is shown as a blue line.
    \figurelabel{transit-ml}}
\end{center}
\end{figure}

>>>>

\begin{figure}[!hptb]
\begin{center}
\includegraphics[width=0.95\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
    $S(d_j)/S(0) = 100$.
    (middle) The corresponding autocorrelation functions with the same colors.
    \response{(right) Three realizations in arbitrary units from each model in
    the same colors.}
    \figurelabel{sho}}
\end{center}
\end{figure}

@@

In any application of \celerite~--~or any other GP model, for that
matter~--~we must choose a kernel.
A detailed discussion of model selection is beyond the scope of this paper and
the interested reader is directed to Chapter~5 of \citet{Rasmussen:2006} for
an introduction in the context of GPs, but we include a summary and some
suggestions specific to \celerite\ here.

@@

Sometimes, it is not possible to choose a physically motivated kernel when we
use a GP as an effective model.
In these cases, we recommend starting with a mixture of stochastically-driven
damped SHOs (as discussed in \sect{sho}) and select the number of oscillators
using Bayesian model comparison, cross-validation, or another model selection
technique.
Despite the fact that it is less flexible than a general \celeriteterm\
kernel, we recommend the SHO model in most cases because it is once mean
square differentiable
\begin{eqnarray}
\left.\frac{\dd k_\mathrm{SHO}(\tau)}{\dd \tau}\right|_{\tau=0} = 0 \quad,
\end{eqnarray}
giving smoother functions than the general \celeriteterm\ model that is only
mean square continuous.

@@

In this first example, we simulate a dataset using a known \celeriteterm\
process and fit it with \celerite\ to show that valid inferences can be made
in this idealized case.
\response{We simulate a small ($N = \exampleindata$) dataset using a GP model
with a SHO kernel (\eqalt{sho-kernel}) with parameters $S_0 =
1\,\mathrm{ppm}^2$, $\omega_0 = e^2\,\mathrm{day}^{-1}$, and $Q = e^2$, where
the units are arbitrarily chosen to simplify the discussion.
We add further white noise with the amplitude $\sigma_n = 2.5\, \mathrm{ppm}$
to each data point.
The simulated data are plotted in the top left panel of
\Figure{simulated-correct}.

@@

\begin{figure}[!hptb]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/simulated/correct.pdf}
\caption{(top left) A simulated dataset (black error bars), and MAP
    model (blue contours).
    (bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
    (right) The inferred PSD~--~the blue contours encompass 68\% of the
    posterior mass~--~compared to the true PSD (dashed black line).
    \figurelabel{simulated-correct}}
\end{center}
\end{figure}

@@

\begin{figure}[!hptb]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/simulated/wrong-qpo.pdf}
\caption{(top left) A simulated dataset (black error bars), and MAP
    model (blue contours).
    (bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
    (right) The inferred period of the process. The true period is indicated
    by the vertical orange line, the posterior inference using the correct
    model is shown as the blue dashed histogram, and the inference made using
    the ``wrong'' effective model is shown as the black histogram.
    \figurelabel{simulated-wrong}}
\end{center}
\end{figure}

\begin{figure}[!hptb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/simulated/bic.pdf}
\caption{\response{
    The Bayesian information criterion (BIC) evaluated for the data from
    \Figure{simulated-wrong} and different kernel choices.
    The $x$-axis shows the number of terms included in each model.
    The blue circles show models where the terms have been summed and the
    orange squares indicate that the model is a product of terms.
    In the product models, the parameters $S_0$ and $Q$ are held fixed at $1$
    and $1/\sqrt{2}$ respectively for all but the first term.
    The dashed black line shows the minium value of the BIC and this
    corresponds to the model chosen in the text.}
    \figurelabel{bic}}
\end{center}
\end{figure}

@@

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data \citep{Angus:2017}.
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P_\mathrm{rot}} \right) \right)
\end{eqnarray}
where $P_\mathrm{rot}$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We construct a simple \celeriteterm\ covariance function with similar
properties as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{B}{2+C}\,e^{-\tau/L}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{rot}}\right) + (1 + C)
\right]
\end{eqnarray}
for $B>0$, $C>0$, and $L>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a significant gain in computational efficiency.

@@

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln
P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
samples.
These samples give a posterior constraint on the period of $P_\mathrm{rot} =
\rotationperiod$ days and the marginalized posterior distribution for $P$ is
shown in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but \citet{Angus:2017} demonstrate
that GP inferences are, at a population level, more reliable than other
methods.

@@

\begin{figure}[hptb]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/rotation/rotation.pdf}
\caption{Inferred constraints on a quasiperiodic GP model using the covariance
    function in \eq{rot-kernel} and two quarters of \kepler\ data.
(top left) The \kepler\ data (black points) and the MAP model
    prediction (blue curve) for a 60~day subset of the data used.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
(right) The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
    figure shows the posterior distribution marginalized over all other
    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
    published rotation period made using the full
    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
    \figurelabel{rotation}}
\end{center}
\end{figure}

@@

\begin{figure}[!hptb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/astero/astero.pdf}
\caption{
\response{
(top) The \kepler\ data (black points) and the MAP model
    prediction (blue curve) for a 10~day subset of the month-long dataset that
    was used for the fit.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom) The residuals between the mean predictive model and the data
    shown in the top figure.
}
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

\begin{figure}[!hptb]
\begin{center}
\includegraphics[width=\textwidth]{figures/transit/transit-ml.pdf}
    \caption{\emph{(top)} A month-long segment of \kepler\ light curve for
    KIC~1430163 with a synthetic transit model injected (black points) and the
    MAP model for the stellar variability
    (blue line).
    \emph{(bottom)} The MAP ``de-trending'' of the data in
    the top panel.
    In this panel, the MAP model for the stellar variability
    has been subtracted to leave only the transits.
    The de-trended light curve is shown by black error bars and the MAP
    transit model is shown as a blue line.
    \figurelabel{transit-ml}}
\end{center}
\end{figure}
